An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-11-01 scagnostics.Rmd
More facets for smoothing compairson.

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the scagnostics on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $p$ is used for the number of points, $n$ is the number of dimensions, and $d=2$ is the projection dimension.

- **pipe**: nuisance directions are generated by sampling independently from a uniform distribution between $[-1,1]$, and the circle is generated by sampling randomly on a 2D circle, and adding a small radial noise.
- **sine**:  nuisance directions are generated by sampling independently from a standard normal distribution, and the sine curve is generated by $x_n = \sin(x_{n-1}) + \mathrm{jittering}$.
- **spiral**:  nuisance directions are generated by sampling independently from a normal distribution, and the structure directions are sampled from an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. The absolute value of $\theta$ fixes the direction of the spiral shape.

@@

To test the behaviour of the index functions we use three types of simulated data sets based on different underlying distributions, and with different built-in features. They are defined in $n=6$ dimensions with $p$ randomly selected points $x_i, i = 1,..,n$ according to:

- spiral:  XXX $p$ points on an n-dimensional sphere randomly selected using the geozoo package [@geozoo], as an example of dataset without special views, labelled "sphere"
- pipe: $p$ points where $n-1$ points are independently drawn from a uniform distribution between $[-1,1]$, and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- sine:  $p$ points where $n-1$ points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"

For each distribution we consider a small sample of $p=100$ and a large sample with $p=1000$.

@@

```{r spiralScagSmoothing, fig.cap="Comparing the traces of the three scagnostics indices when changing the binning via the bins parameter set to 10, 20 and 50 in this example. For too rough binning (bins=10) we observe random fluctuations from the binning, with structure detedction being obscured. With bins=20 we see somewhat smoother behaviour for the stringy index, especially when moving into the maximum, while the skinny index shows a dip in the final view, and larger bins=50 is preferred. For the convex index we also find bins=50 to be the better choice."}

@@

ggplot(scagSmoothMelt, aes(x=t, y=value)) +
  geom_line(aes(color=method, linetype=factor(alpha))) + 
  facet_wrap(~variable, ncol=1, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE))

>>>>

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the scagnostics on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are:

- **pipe**: $n$ points where all $p$ directions are independently drawn from a uniform distribution between $[-1,1]$, but enforcing $x_p^2 + x_{p-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- **sine**:  $n$ points where $p-1$ directions are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_p = \sin(x_{p-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"
- **spiral**:  $n$ points where $p-2$ directions are independently drawn from a normal distribution, and direction $p-1$ and $p$ are sampled on an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. We take the absolute value of $\theta$ to fix the direction of the spiral shape when sampling the points.

@@

@@

```{r spiralScagSmoothing, fig.cap="Comparing the traces of the three scagnostics indices when smoothing the index values, either by averaging over the index value after jittering the projection by some angle alpha (left) or after jittering the projected datapoints with some amount alpha (right). While the behaviour can be somewhat improved, sharp jumps can be observed in all settings considered."}

@@

ggplot(scagSmoothMelt, aes(x=t, y=value)) +
  geom_line(aes(color=factor(alpha))) + 
  facet_wrap(variable~method, ncol=2, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE))
